\chapter{Miniproject: Adeles}\label{Adele_miniproject}

\section{Status}

This is an active miniproject.

\section{The goal}

There are several goals to this miniproject.

\begin{enumerate}
  \item Define the adeles $\A_K$ of a number field~$K$ and
    give them the structure of a $K$-algebra (status: now in mathlib thanks to
    Salvatore Mercuri);
  \item Prove that $\A_K$ is a locally compact topological ring (status:
      \href{https://github.com/smmercuri/adele-ring_locally-compact}{
      also proved by Mercuri} but not yet in mathlib);
  \item Base change: show that if $L/K$ is a finite extension of number fields then the
    natural map $L\otimes_K\A_K\to\A_L$ is an isomorphism, both algebraic and
    topological; (status: not
    formalized yet, but there is a plan -- see the project dashboard);
  \item Prove that $K \subseteq \A_K$ is a discrete subgroup and the quotient
    is compact (status: not formalized yet, but there is a plan -- see the project
    dashboard);
  \item Get this stuff into mathlib (status: (1) done, (2)--(4) not done).
\end{enumerate}

We briefly go through the basic definitions. Let $K$ be a number field.
Let $\Zhat=\projlim_{N\geq1}(\Z/N\Z)$ be the profinite completion of $\Z$,
equipped with the projective limit topology.

A cheap definition of the finite adeles $\A_K^\infty$ of $K$ is $K\otimes_{\Z}\Zhat$,
equipped with the $\Zhat$-module topology.
A cheap definition of the infinite adeles
$K_\infty$ of $K$ is $K\otimes_{\Q}\R$ with the $\R$-module topology (this is a
finite-dimensional $\R$-vector space so this is just the usual topology on $\R^n$).
A cheap definition of the adeles of $K$ is $\A_K^\infty\times K_\infty$ with
the product topology. This is a commutative topological ring.

However in the literature (and in mathlib) we see different definitions.
The finite adeles of $K$ are usually defined in the books
as the so-called restricted product $\prod'_{\mathfrak{p}}K_{\mathfrak{p}}$ over the completions
$K_{\mathfrak{p}}$ of $K$ at all maximal ideals $\mathfrak{p}\subseteq\mathcal{O}_K$ of the
integers of $K$. Here the restricted product is the subset of $\prod_{\mathfrak{p}}K_{\mathfrak{p}}$
consisting of elements which are in the integers $\mathcal{O}_{K,\mathfrak{p}}$ of
$K_{\mathfrak{p}}$ for all but finitely many $\mathfrak{p}$. This is the definition given in
mathlib. Mathlib also has the proof that they're a topological ring;
furthermore the construction of the finite adeles in mathlib works for any
Dedekind domain (this was pointed out to me by Mar\'ia In\'es
de Frutos Fern\'andez; the adeles
are an arithmetic object, but the finite adeles are an algebraic object).

Similarly the infinite adeles of a number field~$K$
are usually defined as $\prod_v K_v$,
the product running over the archimedean completions of~$K$, and this is
the mathlib definition.

The adeles of a number field $K$ are the product of the finite and infinite
adeles, and mathlib knows that they're a $K$-algebra and a topological ring.

\section{Local compactness}

As mentioned above, Salvatore Mercuri was the first to give a complete formalisation of the proof
that the adele ring is locally compact as a topological space. His work is in
\href{https://github.com/smmercuri/adele-ring_locally-compact}{his own repo} and proved the
result using the ``ad hoc'' topology on the adeles which we initially had. Since then,
adeles have been refactored to have the direct limit topology and mathlib has
\href{https://leanprover-community.github.io/mathlib4_docs/Mathlib/Topology/Algebra/RestrictedProduct.html\#RestrictedProduct.locallyCompactSpace_of_addGroup}
{\tt RestrictedProduct.locallyCompactSpace\_of\_addGroup}, the result
that a restricted product of topological additive groups $K_v$ over compact open
subgroups $A_v$ is locally compact.

What we need then is this (note that this is not true for a general Dedekind domain):
\begin{theorem}
  \label{NumberField.instCompactSpaceAdicCompletionIntegers}
  \lean{NumberField.instCompactSpaceAdicCompletionIntegers}
  \discussion{451}
  \leanok
  If $K$ is a number field and $v$ is a nonzero prime ideal of the integers of $K$,
  then the integers of $K_v$ is a compact open subgroup.
\end{theorem}
\begin{proof} Openness should follow from the fact that the integers are
  $\{x : v(x)<v(1/\pi)\}$ where $\pi$ is a uniformizer. Compactness needs
  finiteness of the residue field $\mathcal{O}_K/v$.
\end{proof}

Once we have this, the above result from mathlib gives us
\begin{theorem}
  \lean{NumberField.AdeleRing.locallyCompactSpace}
  \label{NumberField.AdeleRing.locallyCompactSpace}
  \uses{NumberField.instCompactSpaceAdicCompletionIntegers}
  \discussion{253}
  \leanok
  The adeles of a number field are locally compact.
\end{theorem}
\begin{proof}
  The adeles of a number field are a product of the finite adeles and the infinite adeles
  so it suffices to prove that the finite and infinite adeles are locally compact.
  The infinite adeles are just isomorphic to $\R^n$ as a topological space, so they're certainly
  locally compact. As for the finite adeles,
  the mathlib theorem {\tt RestrictedProduct.locallyCompactSpace\_of\_addGroup}
  says that a restricted product of locally compact additive groups with respect to open compact
  subgroups is locally compact, so this reduces us the previous result.
\end{proof}

\section{Base change}

The ``theorem'' we want is that if $L/K$ is a finite extension of number fields,
then $\A_L=L\otimes_K\A_K$. This isn't a theorem though, this is actually a \emph{definition}
(the map between the two objects) and a theorem about
the definition (that it's an isomorphism). In fact the full claim is that it is both a homeomorphism
and an $L$-algebra isomorphism. Before we can prove the theorem, we need to make the
definition.

Recall that the adeles $\A_K$ of a number field is a product $\A_K^\infty\times K_\infty$
of the finite adeles and the infinite adeles. So our ``theorem'' follows immediately from
the ``theorem''s that $\A_L^\infty=L\otimes_K\A_K^\infty$ and $L_\infty=L\otimes_KK_\infty$
(both of these equalities mean an algebraic and topological isomorphism).
We may thus treat the finite and infinite results separately.

\subsection{Base change for nonarchimedean completions.}

As pointed out above, the theory of finite adeles works fine for Dedekind domains.
So for the time being let~$A$ be a Dedekind domain. Recall that the \emph{height one spectrum}
of $A$ is the nonzero prime ideals of~$A$. Note that because we stick to the literature,
rather than to common sense, fields are Dedekind domains in mathlib, and the
height one spectrum of a field is empty. The reason I don't like allowing fields
to be Dedekind domains is that geometrically the definition of Dedekind
domain in the literature is ``smooth affine curve, or a point''. But many theorems in algebraic
geometry begin ``let $C$ be a smooth curve'', rather than ``let $C$ be a smooth curve or a point''.

Let $K$ be the field of fractions of $A$. If $v$ is in the height one spectrum of $A$,
then we can put the $v$-adic topology on $A$ and on $K$, and consider the completions
$A_v$ and $K_v$. The finite adele ring $\mathbb{A}_{A,K}^\infty$ is defined to be
the restricted product of the $K_v$ with respect to the $A_v$, as $v$ runs over
the height one spectrum of $A$. It is topologised by making $\prod_v A_v$ open with
the product topology (here $A_v$ has the $v$-adic topology).

Now let~$L/K$ be a finite separable extension, and let $B$ be the integral closure of~$A$ in~$L$.
We want to relate the finite adeles of $K$ and of $L$. We work place by place, starting by fixing
one place $w$ of $B$ and analysing the relation of $L_w$ and $B_w$ to the completions $K_v$
and $A_v$ where $v$ is the place of $A$ dividing $w$.

So let $w$ be a nonzero prime ideal of $B$. Say $w$ lies over $v$, a prime ideal of $A$.
Then we can put the $w$-adic topology on $L$ and the $v$-adic topology on~$K$. Furthermore
we can equip $K$ with an additive $v$-adic valuation, that is,
a function also called $v$ from $K$ to $\Z\cup\{\infty\}$ normalised so that if $\pi$ is a uniformiser
for $v$ then $v(\pi)=1$. Similarly we consider $w$ as a function from $L$ to $\Z\cup\{\infty\}$.
The next lemma explains how these valuations are related.

\begin{lemma}
  \label{IsDedekindDomain.HeightOneSpectrum.valuation_comap}
  \lean{IsDedekindDomain.HeightOneSpectrum.valuation_comap}
  \leanok
  If $i:K\to L$ denotes the inclusion then for $k\in K$ we have
  $e\times w(i(k))=v(k)$, where $e$ is the ramification index of $w/v$
  (recall that valuations here are written additively, unlike in mathlib).
\end{lemma}
\begin{proof}
  \leanok
  Standard (and formalized).
\end{proof}

\begin{definition}
  \lean{IsDedekindDomain.HeightOneSpectrum.adicCompletionComapSemialgHom}
  \label{IsDedekindDomain.HeightOneSpectrum.adicCompletionComapSemialgHom}
  \uses{IsDedekindDomain.HeightOneSpectrum.valuation_comap}
  \leanok
  There's a natural ring map $K_v\to L_w$ extending the map $K\to L$.
  It is defined by completing
  the inclusion $K\to L$ at the finite places $v$ and $w$ (which can be done
  because the previous lemma shows that the map is uniformly continuous for the $v$-adic
  and $w$-adic topologies).
\end{definition}

\begin{lemma}
  \lean{IsDedekindDomain.HeightOneSpectrum.valued_adicCompletionComap}
  \label{IsDedekindDomain.HeightOneSpectrum.valued_adicCompletionComap}
  \uses{IsDedekindDomain.HeightOneSpectrum.valuation_comap,
  IsDedekindDomain.HeightOneSpectrum.adicCompletionComapSemialgHom}
  \leanok
  If $i_v:K_v\to L_w$ denotes the map of the previous definition
  then for $x\in K_v$ we have
  $e\times w(i(k))=v(k)$, where $e$ is the ramification index of $w/v$.
\end{lemma}
\begin{proof}
  \leanok
  Follows by continuity from lemma~\ref{IsDedekindDomain.HeightOneSpectrum.valuation_comap}.
\end{proof}

\begin{lemma}
  \lean{IsDedekindDomain.HeightOneSpectrum.adicCompletionComapSemialgHom.mapadicCompletionIntegers}
  \label{IsDedekindDomain.HeightOneSpectrum.adicCompletionComapSemialgHom.mapadicCompletionIntegers}
  \uses{IsDedekindDomain.HeightOneSpectrum.valued_adicCompletionComap}
  \leanok
  The map $i_v:K_v\to L_w$ sends the integer ring $A_v$ into $B_w$.
\end{lemma}
\begin{proof}
  \leanok
  The integer ring is defined by $v\geq0$ (or $v\leq 1$ in mathlib, which uses multiplicative
  valuations) so the result follows from \ref{IsDedekindDomain.HeightOneSpectrum.valued_adicCompletionComap}.
\end{proof}
\begin{theorem}
  \lean{IsDedekindDomain.HeightOneSpectrum.adicCompletionComap_isModuleTopology}
  \label{IsDedekindDomain.HeightOneSpectrum.adicCompletionComap_isModuleTopology}
  \uses{IsDedekindDomain.HeightOneSpectrum.adicCompletionComapSemialgHom}
  \leanok
  \discussion{326}
  Giving $L_w$ the $K_v$-algebra structure coming from the natural map $K_v\to L_w$,
  the $w$-adic topology on $L_w$ is the $K_v$-module topology.
\end{theorem}
\begin{proof}
  Any basis for $L$ as a $K$-vector space spans $L_w$ as a $K_v$-module, so $L_w$ is
  finite-dimensional over $K_v$ and the module topology is the same as the product
  topology. So we need to establish that the product topology on $L_w=K_v^n$ is
  the $w$-adic topology. But the $w$-adic topology is induced by the $w$-adic norm,
  which makes $L_w$ into a normed $K_v$-vector space, and (after picking a basis)
  the product norm on $L_w=K_v^n$ also makes $L_w$ into a normed $K_v$-vector space.
  So the result follows from the standard fact (see for example the lemma on p52
  of Cassels-Froelich, formalized as {\tt ContinuousLinearEquiv.ofFinrankEq} in mathlib)
  that any two norms on a finite-dimensional vector space over
  a complete field are equivalent (and thus induce the same topology).
\end{proof}

Because of the commutative diagram
\begin{center}
\begin{tikzcd}
K_v \arrow{r} & L_w  \\
K \arrow{u} \arrow{r} & L \arrow{u}
\end{tikzcd}
\end{center}

we can view $L_w$ as an $L\otimes_KK_v$-algebra.

Now instead of fixing $w$ upstairs, we fix $v$ downstairs and consider all $w$ lying
over it at once. So say $v$ is in the height one spectrum of $A$.

\begin{lemma}
  \lean{IsDedekindDomain.HeightOneSpectrum.Extension.finite}
  \label{IsDedekindDomain.HeightOneSpectrum.Extension.finite}
  \leanok
  There are only finitely many primes $w$ of $B$ lying above $v$.
\end{lemma}
\begin{proof}
  \leanok
  This is a standard fact about Dedekind domains. The key input is
  mathlib's theorem {\tt primesOver\_finite}.
\end{proof}

We write $w|v$ to denote the fact that $w$ is a prime of $B$ above $v$ of $A$.

\begin{definition}
  \lean{IsDedekindDomain.HeightOneSpectrum.adicCompletionComapSemialgHom'}
  \label{IsDedekindDomain.HeightOneSpectrum.adicCompletionComapSemialgHom'}
  \uses{IsDedekindDomain.HeightOneSpectrum.adicCompletionComapSemialgHom}
  \leanok
  The product of the maps $K_v\to L_w$ for $w|v$ is a natural ring map $K_v\to\prod_{w|v}L_w$
  lying over $K\to L$.
\end{definition}

Because $K_v\to\prod_{w|v}L_w$ lies over $K\to L$, there's an induced $L$-algebra
map $L\otimes_KK_v\to\prod_{w|v}L_w$. We are now able to state one of the key results
in this section. The proof is probably the hardest proof
in this section to formalize.

\begin{theorem}
  \lean{IsDedekindDomain.HeightOneSpectrum.adicCompletionComapAlgEquiv}
  \label{IsDedekindDomain.HeightOneSpectrum.adicCompletionComapAlgEquiv}
  \uses{IsDedekindDomain.HeightOneSpectrum.adicCompletionComapSemialgHom',
  IsDedekindDomain.HeightOneSpectrum.Extension.finite}
  \leanok
  The induced $L$-algebra homomorphism $L\otimes_KK_v\to\prod_{w|v}L_w$ is an
  isomorphism of rings.
\end{theorem}
\begin{proof}

  My current proposal to formalize this is as follows. The map is surjective
  because the image is dense and closed; this has been formalized already.
  It is also a $K_v$-algebra homomorphism if we give $L_w$ the obvious $K_v$-algebra
  structure. Thus we can conclude the result if we can prove that both spaces are
  finite-dimensional and have the same dimension. The $K_v$-dimension of $L\otimes_KK_v$
  is equal to the $K$-dimension of $L$, which is $\sum_{w|v}e_wf_w$ using the standard
  notation that $e_w$ is the ramification index of $w$ and $f_w$ the residue degree
  (this result is in mathlib). So it suffices to prove that $[L_w:K_v]=e_wf_w$.
  We already have that $e_w$ (defined globally) is equal to the local ramification
  index (defined as the factor by which the valuations differ on $K$). So what is left
  is to prove that (i) the residue field extension induced by $L_w/K_v$ has degree is equal to the
  globally-defined $f_w$, (ii) an extension of local fields has degree $ef$. Now (i) sounds
  straightforward given what we have (the map from $A$ to $\mathcal{O}_v$ has kernel $v$ and
  dense image) and (ii) is true for any complete discretely-valued field; I am not suggesting
  we formalize the following proof, but at least it represents a rigorous justification:
  A field complete with respect to a discrete valuation is \emph{stable} in the sense
  of the book by Bosch-G\"{u}ntzer-Remmert (Prop 3.6.2.1), so every finite extension of such a field
  is cartesian (def 3.6.1.1) and thus $ef=n$ (Prop 3.6.2.4, (iii) implies (ii)). Note
  that if you weaken the hypotheses too much then there are counterexamples; it's possible
  to have $ef<n$ and BGR goes into details.

\end{proof}

\begin{theorem}
  \label{IsDedekindDomain.HeightOneSpectrum.prodAdicCompletionComap_isModuleTopology}
  \lean{IsDedekindDomain.HeightOneSpectrum.prodAdicCompletionComap_isModuleTopology}
  \uses{IsDedekindDomain.HeightOneSpectrum.adicCompletionComap_isModuleTopology}
  \leanok
  For $v$ fixed, the product topology on $\prod_{w|v}L_w$ is the $K_v$-module
  topology.
\end{theorem}
\begin{proof}
  \leanok
  This is a finite product of $K_v$-modules each of which has the $K_v$-module topology
  by~\ref{IsDedekindDomain.HeightOneSpectrum.adicCompletionComap_isModuleTopology},
  and the product topology is the module topology for a finite product of modules each of which
  has the module topology (this is in mathlib).
\end{proof}

\begin{theorem}
  \label{IsDedekindDomain.HeightOneSpectrum.adicCompletionComapContinuousAlgEquiv}
  \lean{IsDedekindDomain.HeightOneSpectrum.adicCompletionComapContinuousAlgEquiv}
  \uses{IsDedekindDomain.HeightOneSpectrum.prodAdicCompletionComap_isModuleTopology}
  If we give $L\otimes_KK_v$ the $K_v$-module topology then the $L$-algebra isomorphism
  $L\otimes_K K_v\cong\prod_{w|v}L_w$ is also a homeomorphism.
\end{theorem}
\begin{proof} Indeed, is a $K_v$-algebra isomorphism between two modules each of which
  have the module topology, and any module map is automorphically continuous for the
  module topologies.
\end{proof}

We now start thinking about what's going on at the integral level. We write $A_v$
for the integers of $K_v$ and $B_w$ for the integers of $L_w$.

\begin{theorem}
  \lean{IsDedekindDomain.HeightOneSpectrum.adicCompletionComapAlgEquiv_integral}
  \label{IsDedekindDomain.HeightOneSpectrum.adicCompletionComapAlgEquiv_integral}
  \uses{IsDedekindDomain.HeightOneSpectrum.adicCompletionComapAlgEquiv}
  \leanok
  The isomorphism $L\otimes_KK_v\to\prod_{w|v}L_w$ induces an isomorphism
  $B\otimes_AA_v\to \prod_{w|v}B_w$
  for all $v$ in the height one spectrum of $A$.
\end{theorem}
\begin{proof}
  \leanok
  Certainly the image of the integral elements are integral. The argument in the other
  direction is more delicate. My original plan was to follow Cassels--Froehlich,
  Cassels' article ``Global fields'', section 12 lemma, p61, which proves it for
  all but finitely many primes, but \href{https://github.com/ImperialCollegeLondon/FLT/pull/400}
  {a PR by Matthew Jasper} gives another approach which works for all primes.
  Jasper's argument is to show that the closure of $A$ in $K_v$ is $A_v$
  for a valuation on a Dedekind domain, and then that the closure of $A$ in $\prod_{v\in S}K_v$
  is $\prod_{v\in S}A_v$ for $S$ a finite set of valuations (using the Chinese
  remainder theorem). Applying this to $B$ we get that the closure of $B$ in $\prod_{w|v}L_w$
  is $\prod_{w|v}B_w$. He then shows that this closure is the image of
  $B\otimes_A\mathcal{O}_v$ (by showing that this image is closed because it's open),
  giving surjectivity; injectivity follows from the statement
  that $L\otimes_KK_v=\prod_{w|v}L_w$.
\end{proof}

A summary of what we have so far: for all finite places $v$ of $A$
we have shown that the natural map $L\otimes_KK_v\to\prod_wL_w$
is an isomorphism of $L$-algebras, and that if $L\otimes_KK_v$ has
the $K_v$-module topology and each $L_w$ has the valuation topology
then this map is also a homeomorphism. Furthermore we have shown
that there is an induced algebraic isomorphism $B\otimes_AA_v\equiv\prod_w B_w$
on the subrings of the left and right hand sides.

Recall that the finite adeles $\A_{A,K}^\infty$ is defined in mathlib to be
the restricted product of the $K_v$ with respect to the $A_v$, equipped with a certain
restricted product topology (which is not the subspace topology of the product
topology, indeed $\prod_v A_v$ is open in this topology). We have seen in
definition~\ref{IsDedekindDomain.HeightOneSpectrum.adicCompletionComapSemialgHom} that
there's a map $K_v\to L_w$ if $w|v$, extending $K\to L$, and we have seen in
theorem~\ref{IsDedekindDomain.HeightOneSpectrum.adicCompletionComapSemialgHom.mapadicCompletionIntegers}
that this sends $A_v$ to $B_w$. We conclude

\begin{definition}
  \label{IsDedekindDomain.FiniteAdeleRing.mapSemialgHom}
  \lean{IsDedekindDomain.FiniteAdeleRing.mapSemialgHom}
  \uses{IsDedekindDomain.HeightOneSpectrum.adicCompletionComapSemialgHom,
  IsDedekindDomain.HeightOneSpectrum.adicCompletionComapSemialgHom.mapadicCompletionIntegers}
  \leanok
  There's a natural ring homomorphism
  $\A_{A,K}^\infty\to\A_{B,L}^\infty$ lying over $K\to L$.
\end{definition}

Hence there's a natural $L$-algebra homomorphism $L\otimes_K\A_{A,K}^\infty\to\A_{B,L}^\infty$.

Our next goal in this section is the following two results. First the algebraic claim:

\begin{theorem}
  \label{IsDedekindDomain.FiniteAdeleRing.baseChangeAlgEquiv}
  \lean{IsDedekindDomain.FiniteAdeleRing.baseChangeAlgEquiv}
  \uses{IsDedekindDomain.HeightOneSpectrum.adicCompletionComapAlgEquiv_integral,
    RestrictedProduct.relabelIso,
    IsDedekindDomain.FiniteAdeleRing.baseChangeIntegralAlgEquiv,
    IsDedekindDomain.AKLB.tensorProduct_module_algEquiv}
  \leanok
  \discussion{243}
  This natural map $L\otimes_K\A_{A,K}^\infty\to\A_{B,L}^\infty$ is an isomorphism.
\end{theorem}

Now $L\otimes_K\A_{A,K}^\infty$ is an $\A_{A,K}^\infty$-module and hence can be given
the $\A_{A,K}^\infty$-module topology. We also claim

\begin{theorem}
  \label{IsDedekindDomain.FiniteAdeleRing.baseChangeContinuousAlgEquiv}
  \lean{IsDedekindDomain.FiniteAdeleRing.baseChangeContinuousAlgEquiv}
  \uses{IsDedekindDomain.HeightOneSpectrum.adicCompletionComapAlgEquiv_integral}
  \leanok
The induced $L$-algebra morphism
  $L\otimes_K\A_{A,K}^\infty\to\A_{B,L}^\infty$ is a topological isomorphism.
\end{theorem}

Informally, the proofs are simple: componentwise we know
that $L\otimes_KK_v$ is isomorphic both algebraically and
topologically to $\prod_{w|v}L_w$, and that this isomorphism
sends the open set $B\otimes_AA_v$ homeomorphically onto
$\prod_{w|v}B_w$, so now it's ``just a case of putting everything
together''. Formally, we really need to spell this out, as there is a lot
going on. We do this in the
next subsection.

\subsection{Base change for nonarchimedean completions.}

As usual we are in the AKLB set-up, so in particular $K$ is the field
of fractions of the Dedekind domain $A$, $L/K$ is a finite
separable extension, and $B$ is the integral closure of $A$
in $L$. The goal in this subsection is to spell out the following argument: Assume that
$L\otimes_KK_v\cong\prod_{w|v}L_w$
algebraically and topologically for all $v$, with $B\otimes_AA_v$ identified with $\prod_{w|v}B_w$.
Then $L\otimes_K\A_K^\infty\cong\A_L^\infty$, algebraically and topologically. Here
the tensor products $L\otimes_K R$ (for $R$ a $K$-algebra with a topology) are all being
given the $R$-module topology, which if we choose a basis for $L/K$ is just the product
topology.

We start with the following observation. If $M$ is a $K$-module then there's a canonical
map $B\otimes_AM\to L\otimes_KM$ sending $b\otimes m$ to $b\otimes m$ (this follows from the
universal property of the tensor product). Our first goal
is to show that this map is an isomorphism. Let us establish some lemmas first.

\begin{lemma}
  \label{IsDedekindDomain.dvd_norm}
  If $0\not=b\in B$ then there exists $0\not=a\in A$ such that $b$ divides
  the image of $a$ in $B$.
\end{lemma}
\begin{remark} Is this already in mathlib?
\end{remark}
\begin{proof} Let $a=N_{L/K}(b)$, the norm. This is known to take nonzero elements of $L$
to nonzero elements of $K$ (because the norm is the determinant of an invertible linear map)
and integral elements to integral elements. Furthermore $a/b\in L$ is the the product of the
conjugates of $b$ in some normal closure of $L$, and hence it is integral, and thus in $B$.
\end{proof}

\begin{corollary}
  \label{IsDedekindDomain.AKLB.surjective_tensorProduct_map}
  \uses{IsDedekindDomain.dvd_norm}
  The $A$-bilinear map $B\times K\to L$ sending $(b,k)$ to $bk$ is surjective.
\end{corollary}
\begin{proof} Given $\lambda\in L$ write it as $n/d$ with $0\not=d\in B$. Choose $0\not=a\in A$
  and $b\in B$ with $db=a$ and then note $\lambda=nb/a=nb\times a^{-1}$.
\end{proof}

\begin{corollary}
  \label{IsDedekindDomain.FiniteAdeleRing.tensorProduct_algEquiv}
  \uses{IsDedekindDomain.dvd_norm}
  The natural map $B\otimes_AK\to L$ is a $B$-algebra isomorphism.
\end{corollary}
\begin{proof}

We write down an inverse. Regard $B\otimes_AK$ as a $B$-algebra via the action on the left.
Note that at this point it's not even clear that $B\otimes_AK$ is a field. We have the
structure map $B\to B\otimes_AK$ sending $b$ to $b\otimes1$, which is $B$-linear. I claim
that every nonzero element of $B$ gets sent to an invertible element of $B\otimes_AK$.
Indeed, if $b\not=0$ and (using the previous lemma) we choose $0\not=a\in A$ such that
$bb'=a$, then $(b\otimes1)(b'\otimes\frac1a)=1$. Thus by the universal property of
localisation, the $B$-linear map $B\to B\otimes_AK$ extends to a ring homomorphism from the field
of fractions of $B$ to $B\otimes_AK$, which we claim is our desired inverse.
Checking that both composites are the identity should be straightforward. Starting
with $B\otimes_AK$ we only have to check on elements of the form $b\otimes k$;
starting with $L$ we only have to check on elements of $B$. Hopefully both are
straightforward.
\end{proof}

\begin{corollary}
  \label{IsDedekindDomain.AKLB.tensorProduct_module_algEquiv}
  \uses{IsDedekindDomain.FiniteAdeleRing.baseChangeAlgEquiv}
  If $M$ is any $K$-module then the canonical map $B\otimes_A M\to L\otimes_K M$
  is an isomorphism.
\end{corollary}
\begin{proof} We can factor this map as $B\otimes_AM\cong B\otimes_A(K\otimes_KM)\cong
  (B\otimes_A K)\cong_KM\to L\otimes_KM$ and we just showed that the latter map was an
  isomorphism.
\end{proof}

We now need to explain how tensor products sometimes commute with restricted products.
Something we will need along the way is

\begin{theorem} $B$ is a finitely-presented $A$-module.
  \label{IsDedekindDomain.AKLB.finitePresentation}
\end{theorem}
\begin{proof} $A$ is Noetherian as it is a Dedekind domain, so it suffices to prove that $B$ is
  finitely-generated as an $A$-module. But this is in mathlib already (a proof is
  around line 101 of {\tt BaseChange.lean} in FLT at the time of writing).
\end{proof}

The reason we care about this is the following.

\begin{theorem}
  \label{pi_tensorProduct_of_finitePresentation}
  If $R$ is a commutative ring, if $M$ is a finitely presented $R$-module
  and if $N_i$ are a collection of $R$-modules, then the canonical map
  $M\otimes_R\prod_i N_i\to\prod_i(M\otimes_R N_i)$ is an isomorphism.
\end{theorem}
\begin{proof} If $M$ is finite and free then Maddy Crim has already formalized this
  in FLT. For the general case present $M$ as $R^a\to R^b\to M\to 0$ and use that tensor
  products and arbitrary products preserve surjections.
\end{proof}

\begin{corollary}
  \label{IsDedekindDomain.pi_tensorProduct}
  \uses{IsDedekindDomain.AKLB.finitePresentation,pi_tensorProduct_of_finitePresentation}
  If $S$ is a finite set of nonzero primes of $A$ then the natural map
  $B\otimes((\prod_{v\in S}K_v)\times(\prod_{v\notin S}A_v))\to
  (\prod_{v\in S}(B\otimes_AK_v))\times(\prod_{v\notin S}(B\otimes_AA_v))$
  is an isomorphism.
\end{corollary}
\begin{proof} Follows from the previous two theorems.
\end{proof}

Recall that $\A_K^\infty$ is the finite adeles of $K$,
defined as the restricted product of the $K_v$ with respect to the $A_v$,
where $v$ runs through the nonzero primes of $A$. Let $R$ denote the restricted
product of the $B\otimes_A K_v$ with respect to the $B\otimes_A A_v$.

\begin{corollary}
  \label{IsDedekindDomain.FiniteAdeleRing.IntegraltensorProductAlgEquiv_aux1}
  \uses{IsDedekindDomain.pi_tensorProduct}
  The natural map $B\otimes_A\A_K^\infty\to R$ is a $B$-algebra isomorphism.
\end{corollary}
\begin{proof} This follows from the previous corollary and the fact that
  tensor products commute with filtered colimits.
\end{proof}

Recall from earlier in this section that if $v$ is a finite place of $A$ then the natural map from
$B\otimes_A K_v$ to $L\otimes_KK_v$ is an isomorphism, and recall from the previous section
that the natural map from $L\otimes_KK_v$ to $\prod_{w|v}L_w$ was also an isomorphism.
This isomorphism sends $B\otimes_A A_v$ to $\prod_{w|v}B_w$ (I thank Matthew Jasper for
pointing out to me that this statement was true at all primes, not just at unramified primes).
Finally, the set of $w$ of $B$ dividing a fixed place $v$ of $A$ is finite.
Let's now formalize the abstract statement which we need. Rather than following the
notation for restricted product in the literature and writing $\A_K^\infty=\prod'_vK_v$
with the $\mathcal{O}_v$ implicit, we will write $\prod'_v(K_v,\mathcal{O}_v)$ in the below.

\begin{definition}
  \label{RestrictedProduct.relabelIso}
  Let $V$ and $W$ be index sets, and let $f:W\to V$ be a map with finite fibres.
  Let $X_v$ be sets, with subsets $C_v$, let $Y_w$ be sets with subsets $D_w$,
  and say for all $v\in V$ we're given a bijection $X_v\to\prod_{w|f(w)=v}Y_w$,
  identifying $C_v$ with $\prod_{w:f(w)=v}D_w$. Then there's an induced bijection between
  the restricted products $\prod'_v(X_v,C_v)$ and $\prod'_w(Y_w,D_w)$.
\end{definition}

\begin{corollary}
  \label{IsDedekindDomain.FiniteAdeleRing.IntegraltensorProductAlgEquiv_aux2}
  \uses{RestrictedProduct.relabelIso}
  The ring $R$ introduced above (the restricted
  product of the $B\otimes_A K_v$ with respect to the $B\otimes_A A_v$)
  is isomorphic to $\mathbb{A}_L$.
\end{corollary}
\begin{proof} Let $V$ be the finite places of $K$ and $W$ the finite places of $L$,
  let $X_v$ be $B\otimes_A K_v$, let $C_v$ be $B\otimes_A A_v$, let $Y_w$ be $L_w$,
  let $D_w$ be $B_w$ and the result follows from the previous definition, given
  theorem~\ref{IsDedekindDomain.HeightOneSpectrum.adicCompletionComapAlgEquiv_integral}.
\end{proof}
From this, we can deduce the theorem we claimed earlier:

\begin{theorem}
  \label{IsDedekindDomain.FiniteAdeleRing.baseChangeIntegralAlgEquiv}
  \uses{RestrictedProduct.relabelIso,
  IsDedekindDomain.FiniteAdeleRing.IntegraltensorProductAlgEquiv_aux1,
  IsDedekindDomain.FiniteAdeleRing.IntegraltensorProductAlgEquiv_aux2}
  The natural map $B\otimes_A\A_K^\infty\to\A_L^\infty$ is a $B$-algebra
  isomorphism.
\end{theorem}
\begin{proof}
  This map factors through the auxiliary ring~$R$ so the result follows
  from the previous two constructions.
\end{proof}

Because this map factors through the isomorphism $B\otimes_A\A_K^\infty\to L\otimes_K\A_K^\infty$
we can finally deduce that the natural map $L\otimes_K\A_K^\infty\to\A_L^\infty$ is an algebraic
  isomorphism.
\begin{proof}
  \proves{IsDedekindDomain.FiniteAdeleRing.baseChangeAlgEquiv}
  Follows immediately from theorem~\ref{IsDedekindDomain.FiniteAdeleRing.baseChangeIntegralAlgEquiv}
  and theorem~\ref{IsDedekindDomain.AKLB.tensorProduct_module_algEquiv}.
\end{proof}

We still need to talk about topologies though, so let's finish by doing this. Let's start with some trivialities, expressed as definitions rather than theorems
because they're constructions.

\begin{definition} If $X_v$ and $Y_v$ are families of topological spaces indexed by $v\in V$,
  if $f_v:X_v\to Y_v$ is a continuous map sending the subset $C_v\subseteq X_v$ into
  $D_v\subseteq Y_v$ then there's an induced continuous map $\prod'_v(X_v,C_v)\to\prod'_v(Y_v,D_v)$.
\end{definition}

\begin{definition} If all the $f_v$ are homeomorphisms identifying $C_v$ with $D_v$ then
  the induced map on restricted products is also a homeomorphism (proof: apply the previous
  construction to $f_v$ and their inverses)
\end{definition}

We now allow a slight change of index set. Unfortunately I don't think that we can deduce
the results just stated above from this one, in Lean, because the product of $Y_w$ over a
set of size 1 is not definitionally equal to $Y_w$.

Recall definition~\ref{RestrictedProduct.relabelIso}, giving us a bijection between two restricted
products.

\begin{theorem} In the same setup as definition~\ref{RestrictedProduct.relabelIso}
  ($V,W$ index sets, $f:W\to V$,
  $C_v\subseteq X_v$ and $D_w\subseteq Y_w$, bijections $b_v:X_v\to\prod_{w:f(w)=v}Y_w$
  identifying $C_v$ with $\prod_{w:f(w)=v}D_w$), if all the $X_v$ and $Y_w$ are furthermore
  topological spaces, all the $C_v$ and $D_w$ are open, and all the $b_v$ are homeomorphisms,
  then the induced
  map $\prod'_v(X_v,C_v)\to\prod'_w(Y_w,D_w)$ is also a homeomorphism.
\end{theorem}
\begin{proof} I have only thought about the cofinite filter case, where this
  should follow easily from the definition of the topology.
\end{proof}

\begin{corollary} $\mathbb{\A_L^\infty}$ is homeomorphic to $\prod_v(B\otimes_AK_v,B\otimes_AA_v)$.
\end{corollary}
\begin{proof} Follows from the previous theorem with $X_v=B\otimes_AK_v$ $D_w=L_w$ etc.
\end{proof}

Recall that if $$R$$ is a commutative ring, and two $$R$$-modules both have the $$R$$-module
topology, then any $$R$$-linear morphism between them is automatically continuous. We know
that $\A_L^\infty$ is $\A_K^\infty$-linearly isomorphic to $L\otimes_K\A_K^\infty$ and our claim is that
if $L\otimes_K\A_K^\infty$ is given the $\A_K^\infty$-module topology then this isomorphism is also a
homeomorphism; to prove this, we thus just need to check that $\A_L^\infty$ has the $\A_K^\infty$-module
topology. Equivalently, by the previous result, we need to check that
the restricted product topology on the $\A_K^\infty$-algebra $\prod'_v(B\otimes_AK_v,B\otimes_AA_v)$
is the $\A_K^\infty$-module topology.

We now need to make restricted products of modules into modules over restricted product of rings.
The API, which should be straightforward so we don't give details here, is: if $R_v$ are rings
with subrings $S_v$ and if $M_v$ are $R_v$-modules with $S_v$-stable submodules $N_v$, then
$\prod'_v(M_v,N_v)$ is naturally a module over $\prod'_v(R_v,S_v)$, and that $R_v$-morphisms
$M_v\to M_v'$ sending $N_v$ to $N_v'$ induce $\prod'_v(R_s,S_v)$-linear maps
$\prod'_v(M_v,N_v)\to\prod'_v(M'_v,N'_v)$. From this one deduces that isomorphisms on the factors
induce isomorphisms on the restricted products.


Now $A_v$ is a PID, so $B\otimes_AA_v$ is free (as it is finitely-generated and torsion-free).
This means that there is an isomorphism $B\otimes_AA_v\cong(A_v)^n$, which extends to an isomorphism
$B\otimes_AK_v\cong K_v^n$. These isomorphisms are also homeomorphisms. If we fix such isomorphisms
for all $v$ then we get an induced $\A_K^\infty$-module isomorphism + homeomorphism
$\prod'_v(B\otimes_AK_v,B\otimes_AA_v)=\prod'_v(K_n^n,A_v^n)$. So it suffices to prove
that the $\prod'_v(K_v,A_v)$-module $\prod'_v(K_v^n,A_v^n)$ has the $\prod'_v(K_v,A_v)$-module
topology. This follows from the fact that the product topology on two modules with the
module topology is the module topology (a fact in mathlib) and the following result.

\begin{lemma}
  If $X_v$ and $Y_v$ are topological spaces with open subspaces $C_v$ and $D_v$, then
  the obvious bijection $\prod'_v(X_v \times Y_v,C_v\times D_v) \cong
  \left(\prod'_v(X_v,C_v)\right)\times\left(\prod'_v(Y_v,D_v)\right)$ is a homeomorphism,
  where the restricted products have the restricted product topology and the binary
  product has the product topology.
\end{lemma}
\begin{proof} This should hopefully be straightforward using {\tt RestrictedProduct.continuous\_dom\_prod}
\end{proof}

As a corollary one can prove by induction on $n$ that the restricted product of $n$th powers
is homeomorphic to the $n$th power of the restricted product and this is the result
we require.

\subsection{Base change for infinite adeles}

Recall that if $K$ is a number field then the infinite adeles of $K$ are defined
to be the product $\prod_{v\mid\infty} K_v$ of all the completions of $K$ at the
infinite places.

The result we need here is that if $L/K$ is a finite extension of number fields,
then the map $K\to L$ extends to a continuous $K$-algebra map $K_\infty\to L_\infty$,
and thus to a continuous $L$-algebra isomorphism $L\otimes_KK_\infty\to L_\infty$.
Perhaps a cheap proof would be to deduce it from the fact that $K_\infty=K\otimes_{\Q}\R$.

The overall strategy is to first establish, for each infinite place $v$ of $K$, homeomorphisms
between for the completion $K_v$ and the product $\prod_{w\mid v}L_w$ of completions of $L$ at
all infinite places $w$ of $L$ lying above $v$.
We then use these homeomorphisms to construct base change for the infinite adele ring.

\subsubsection{Weak approximation at infinite places}

First, we require a preliminary result that $K$ is dense inside any product of
completions $\prod_{v\in S} K_v$ of $K$ at infinite places.

\begin{theorem}
  \label{NumberField.InfinitePlace.Completion.denseRange_algebraMap_subtype_pi}
  \lean{NumberField.InfinitePlace.Completion.denseRange_algebraMap_subtype_pi}
  \leanok
  Let $S$ be a set of infinite places of $K$. The image of $K$ under the embedding
  $K\hookrightarrow (K_v)_{v\in S}; k \mapsto (k)_v$ is dense in the product topology.
\end{theorem}
\begin{proof}
  Let $(K, v)$ denote $K$ equipped with the topology induced by the infinite place $v$.
  It suffices to show that the image of $K$ under the embedding
  $K\hookrightarrow \prod_{v\mid \infty}(K, v)$ is dense in the product topology.
  By a standard analytic argument, for each $v$ it is possible to select a sequence $(x_n^{(v)})_n$
  with the property that $x_n^{(v)} \to 1$ in $v$'s topology, while $x_n^{(v)} \to 0$ in any other
  infinite place's topology.
  Let $(z_v)_v \in \prod_{v\mid \infty}(K, v)$.
  For each $v$, the sequence $y_n := \sum_{v\mid\infty} x_n^{(v)}z_v$ in $K$ converges to
  $z_v$ in $v$'s topology.
  So the embedded image of $y_n$ in $\prod_{v\mid \infty}(K, v)$ converges to $(z_v)_v$ in the
  product topology.
\end{proof}

\subsubsection{Dimensionality of $\prod_{w\mid v}L_w$ as a $K_v$-vector space}

This subsection contains a result that the $K_v$-dimension of
$L \otimes_K K_v$ is equal to the $K_v$-dimension of $\prod_{w\mid v}L_w$.

\begin{theorem}
  \label{NumberField.InfinitePlace.Completion.finrank_pi_eq_finrank_tensorProduct}
  \lean{NumberField.InfinitePlace.Completion.finrank_pi_eq_finrank_tensorProduct}
  For a fixed infinite place $v$ of $K$, we have
  $\text{dim}_{K_v} \prod_{w\mid v} L_w = \text{dim}_{K_v} L\otimes_K K_v$.
\end{theorem}

\subsubsection{Base change at infinite places}

\begin{definition}
  \label{NumberField.InfinitePlace.Completion.piExtension}
  \lean{NumberField.InfinitePlace.Completion.piExtension}
  \leanok
  Let $v$ be an infinite place of $K$. There is a continuous $K$-algebra homomorphism
  $K_v \to \prod_{w\mid v}L_w$, whose restriction to $K$ corresponds to the global embedding
  of $K$ into $(L_w)_w$.
\end{definition}

The map in~\ref{NumberField.InfinitePlace.Completion.piExtension} can be lifted to an
$L$-algebra homomorphism defined on $L\otimes_K K_v$.

\begin{definition}
  \label{NumberField.InfinitePlace.Completion.baseChange}
  \lean{NumberField.InfinitePlace.Completion.baseChange}
  \leanok
  \uses{NumberField.InfinitePlace.Completion.piExtension}
  Let $v$ be an infinite place of $K$. There is a natural $L$-algebra homomorphism
  $L\otimes_K K_v \to \prod_{w\mid v}L_w$, whose restriction to $1\otimes_K K_v$ corresponds to
  the map in~\ref{NumberField.InfinitePlace.Completion.piExtension}.
\end{definition}

\begin{theorem}
  \label{NumberField.InfinitePlace.Completion.baseChange_surjective}
  \lean{NumberField.InfinitePlace.Completion.baseChange_surjective}
  \leanok
  \uses{NumberField.InfinitePlace.Completion.denseRange_algebraMap_subtype_pi}
  For a fixed infinite place $v$ of $K$, the map $L\otimes_K K_v \to\prod_{w\mid v}L_w$ is
  surjective.
\end{theorem}
\begin{proof}
  Let $(x_i)_i$ be a $K_v$-basis of $\prod_{w\mid v}L_w$. By the density of $L$ in
  $\prod_{w\mid v}L_w$
  (Theorem~\ref{NumberField.InfinitePlace.Completion.denseRange_algebraMap_subtype_pi}), we can
  find $\alpha_i \in L$ arbitrarily close to $x_i\prod_{w\mid v}L_w$ with respect to the sup norm
  when embedded globally in $\prod_{w\mid v}L_w$.
  In particular, it is possible to choose such $\alpha_i$ so that the matrix representing
  the vector $((\alpha_i)_{w \mid v})_i$ in the basis $(x_i)_i$ has non-zero determinant.
  Since $(\alpha_i)_{w \mid v}$ is the image of $1\otimes \alpha_i$ under base change, we have
  that $(1 \otimes \alpha_i)_i$ also forms a basis of $L\otimes_K K_v$, and base change
  is surjective.
\end{proof}

\begin{theorem}
  \label{NumberField.InfinitePlace.Completion.baseChange_injective}
  \lean{NumberField.InfinitePlace.Completion.baseChange_injective}
  \leanok
  \uses{NumberField.InfinitePlace.Completion.finrank_pi_eq_finrank_tensorProduct}
  For a fixed infinite place $v$ of $K$, the map $L\otimes_K K_v \to\prod_{w\mid v}L_w$ is
  injective.
\end{theorem}
\begin{proof}
  The $L$-algebra map $L\otimes_K K_v \to\prod_{w\mid v}L_w$ can equivalently be thought of
  as $K_v$-linear, which is injective, because it is surjective by
  Theorem~\ref{NumberField.InfinitePlace.baseChange_surjective}, and both side have the same
  $K_v$-dimension by
  Theorem~\ref{NumberField.InfinitePlace.Completion.finrank_pi_eq_finrank_tensorProduct}.
\end{proof}

We have established that the map of
Definition~\ref{NumberField.InfinitePlace.Completion.baseChange} gives an $L$-algebra isomorphism
between $L\otimes_K K_v$ and $\prod_{w\mid v}L_w$.
The left-hand side is given the $K_v$-module topology, while we show that the right-hand side also
has the $K_v$-module topology.
\begin{theorem}
  \label{NumberField.InfinitePlace.Completion.instIsModuleTopologyValEqComapAlgebraMap_fLT}
  \lean{NumberField.InfinitePlace.Completion.instIsModuleTopologyValEqComapAlgebraMap_fLT}
  \leanok
  If $w \mid v$ is an infinite place of $L$ lying above the infinite place $v$ of $K$, then
  $L_w$ has the $K_v$-module topology.
\end{theorem}
\begin{proof}
  Because $L_w$ is a finite-dimensional normed $K_v$ vector space, there exists a $K_v$-linear
  linear homeomorphism $L_w \cong K_v^n$, from which $L_w$ has the $K_v$-module topology.
\end{proof}

\begin{theorem}
  \label{NumberField.InfinitePlace.Completion.baseChangeEquiv}
  \lean{NumberField.InfinitePlace.Completion.baseChangeEquiv}
  \leanok
  \uses{NumberField.InfinitePlace.Completion.baseChange,
    NumberField.InfinitePlace.Completion.baseChange_surjective,
    NumberField.InfinitePlace.Completion.baseChange_injective,
    NumberField.InfinitePlace.Completion.instIsModuleTopologyValEqComapAlgebraMap_fLT}
  Let $v$ be an infinite place of $K$. There is a natural $L$-algebra homeomorphism
  $L\otimes_K K_v \cong_L \prod_{w\mid v}L_w$, whose restriction to $1\otimes_K K_v$ corresponds to
  the map in~\ref{NumberField.InfinitePlace.Completion.piExtension}.
\end{theorem}
\begin{proof}
  The map in~\ref{NumberField.InfinitePlace.Completion.baseChange} is an $L$-algebra
  isomorphism by Theorem~\ref{NumberField.InfinitePlace.Completion.baseChange_surjective}
  and Theorem~\ref{NumberField.InfinitePlace.Completion.baseChange_injective}.
  Every $K_v$-algebra isomorphism between two $K_v$-module topological spaces is a homeomorphism.
  Since the $L$-algebra isomorphism of
  Definition~\ref{NumberField.InfinitePlace.Completion.baseChange} can equivalently be viewed as
  a $K_v$-algebra isomorphism, it is also a homeomorphism.
\end{proof}

\begin{theorem}
  \label{NumberField.InfinitePlace.Completion.piEquiv}
  \lean{NumberField.InfinitePlace.Completion.piEquiv}
  \leanok
  \uses{NumberField.InfinitePlace.Completion.baseChangeEquiv}
  Let $v$ be an infinite place of $K$. There is a natural $K_v$-linear homeomorphism
  $K_v^{[L:K]} \cong_{K_v} \prod_{w\mid v}L_w$.
\end{theorem}
\begin{proof}
  Compose the $K_v$-linear isomorphism $K_v^{[L:K]} \cong \prod_{w\mid v}L_w$ with the $K_v$-linear
  version of base change~\ref{NumberField.InfinitePlace.Completion.baseChangeEquiv} to get the
  isomorphism in the statement.
  Since both sides have the $K_v$-module topology, then this is also a homeomorphism.
\end{proof}

\subsubsection{Base change for the infinite adele ring}

First, we induce a $K_{\infty}$-algebra on $L_{\infty}$ from the action of each $K_v$ on
$\prod_{w\mid v}L_w$.
Specifically, this means that for $x \in K_{\infty}$ and $y \in L_{\infty}$, we have
$(x \cdot y)_w = x_{v_w} \cdot y_w$, where $v_w$ is the restriction of $w$ to $K$.
We show that $L_{\infty}$ has the $K_{\infty}$-module topology.

\begin{theorem}
  \label{NumberField.InfiniteAdeleRing.piEquiv}
  \lean{NumberField.InfiniteAdeleRing.piEquiv}
  \leanok
  \uses{NumberField.InfinitePlace.Completion.piEquiv}
  There is a natural $K_{\infty}$-linear homeomorphism
  $K_{\infty}^{[L:K]} \cong_{K_{\infty}} L_{\infty}$.
\end{theorem}
\begin{proof}
  Using the isomorphisms $K_v^{[L:K]} \cong_{K_v} \prod_{w\mid v}L_w$ from
  Theorem~\ref{NumberField.InfinitePlace.Completion.piEquiv}, we clearly have a bijection
  $K_{\infty}^{[L:K]} \cong \prod_v\prod_{w \mid v} L_w \cong \prod_w L_w$.
  The $K_v$-linearity of each component isomorphism extends to $K_{\infty}$-linearity if the
  action of $\prod_v K_v$ on $\prod_w L_w$ is constant on the fibers of the restriction map on
  infinite places.
  In other words, if, for all $x \in K_{\infty}$ and $y \in L_{\infty}$, we have
  $(x \cdot y)_w = x_{v_w} \cdot y_w$, which is true by definition.
\end{proof}

\begin{theorem}
  \label{NumberField.InfiniteAdeleRing.instIsModuleTopology_fLT}
  \lean{NumberField.InfiniteAdeleRing.instIsModuleTopology_fLT}
  \leanok
  \uses{NumberField.InfiniteAdeleRing.piEquiv}
  $L_{\infty}$ has the $K_{\infty}$-module topology.
\end{theorem}
\begin{proof}
  Since $L_{\infty}$ is homeomorphic to a finite product of $K_{\infty}$ as a $K_{\infty}$-vector
  space, it has the $K_{\infty}$-module topology.
\end{proof}

\begin{theorem}
  \label{NumberField.InfiniteAdeleRing.baseChangeEquivAux}
  \lean{NumberField.InfiniteAdeleRing.baseChangeEquivAux}
  \leanok
  \uses{NumberField.InfinitePlace.Completion.baseChangeEquiv}
  There is a natural $L$-algebra isomorphism
  $L \otimes_K K_{\infty} \cong_L L_{\infty}$.
\end{theorem}
\begin{proof}
  This follows from the following chain of isomorphisms:
  \[
    L \otimes_K K_{\infty} \cong_L \prod_v (L \otimes_K K_v) \cong_L
      \prod_v \prod_{w\mid v}L_w \cong_L L_{\infty} .
  \]
  The first isomorphism is the standard $L$-algebra isomorphism
  $L \otimes_K \prod_v K_v \cong_L \prod_v (L \otimes_K K_v)$.
  The second isomorphism is given by the component $L$-algebra isomorphisms
  $L \otimes_K K_v \cong_L \prod_{w\mid v}L_w$ from
  Theorem~\ref{NumberField.InfinitePlace.Completion.baseChangeEquiv}.
\end{proof}

It remains to show that the map in~\ref{NumberField.InfiniteAdeleRing.baseChangeEquivAux} is a
homeomorphism.
Since both sides have the $K_{\infty}$-module topology, and since the $L$-algebra isomorphism
of~\ref{NumberField.InfiniteAdeleRing.baseChangeEquivAux} can equivalently be viewed as a
$K_{\infty}$-linear isomorphism, it is also a homeomorphism.

\begin{theorem}
  \label{NumberField.InfiniteAdeleRing.baseChangeEquiv}
  \lean{NumberField.InfiniteAdeleRing.baseChangeEquiv,
    NumberField.InfiniteAdeleRing.instIsModuleTopology_fLT}
  \leanok
  \uses{NumberField.InfiniteAdeleRing.baseChangeEquivAux}
  If $K\to L$ is a ring homomorphism between two number fields then there is a natural isomorphism
  (both topological and algebraic) $L\otimes_KK_\infty\cong L_\infty$.
\end{theorem}
\begin{proof}
  Since both sides of the $L$-algebra isomorphism
  in~\ref{NumberField.InfiniteAdeleRing.baseChangeEquivAux} have the $K_{\infty}$-module topology,
  and since the isomorphism can equivalently be viewed as a $K_{\infty}$-linear isomorphism,
  it is also a homeomorphism.
\end{proof}


\subsection{Base change for adeles}

From the previous results we deduce immediately that if $L/K$ is a finite extension
of number fields then there's a natural (topological and algebraic) isomorphism
$L\otimes_K\A_K\to \A_L$.

\begin{theorem}
  \label{NumberField.AdeleRing.baseChangeEquiv}
  \lean{NumberField.AdeleRing.baseChangeEquiv}
  \uses{NumberField.InfiniteAdeleRing.baseChangeEquiv,
  IsDedekindDomain.FiniteAdeleRing.baseChangeAlgEquiv}
  If $K\to L$ is a ring homomorphism between two number fields then there is a natural isomorphism
  (both topological and algebraic) $L\otimes_K\A_K\cong\A_L$.
\end{theorem}
\begin{proof}
  Follows from the previous results.
\end{proof}

Something else we shall need:

\begin{theorem}
  \label{NumberField.AdeleRing.baseChange_moduleTopology}
  \uses{NumberField.AdeleRing.baseChangeEquiv}
  If $K\to L$ is a ring homomorphism between two number fields then the topology on $\A_L$
  is the $\A_K$-module topology, where the module structure comes from the
  natural map $\A_K\to\A_L$.
\end{theorem}
\begin{proof}
  Indeed $\A_L\cong L\otimes_K\A_K$ is a homeomorphism, and
  the right hand side has the $\A_K$-module topology.
\end{proof}

\section{Discreteness and compactness}

We need that if $K$ is a number field then
$K\subseteq\mathbb{A}_K$ is discrete, and the quotient (with the
quotient topology) is compact. Here is a proposed proof.

\begin{theorem}
  \lean{Rat.AdeleRing.zero_discrete}
  \label{Rat.AdeleRing.zero_discrete}
  \leanok
  There's an open subset of $\A_{\Q}$ whose intersection with $\Q$ is $\{0\}$.
\end{theorem}
\begin{proof}
  Use $\prod_p{\Z_p}\times(-1,1)$. Any rational $q$ in this set is a $p$-adic
  integer for all primes $p$ and hence (writing it in lowest terms as $q=n/d$)
  satisfies $p\nmid d$, meaning that $d=\pm1$ and thus $q\in\Z$. The fact
  that $q\in(-1,1)$ implies $q=0$.
\end{proof}

\begin{theorem}
  \lean{NumberField.AdeleRing.zero_discrete}
  \label{NumberField.AdeleRing.zero_discrete}
  \uses{Rat.AdeleRing.zero_discrete,NumberField.AdeleRing.baseChangeEquiv}
  \leanok
  There's an open subset of $\A_{K}$ whose intersection with $K$ is $\{0\}$.
\end{theorem}
\begin{proof}
  By a previous result, we have $\A_K=K\otimes_{\Q}\A_{\Q}$.
  Choose a basis of $K/\Q$; then $K$ can be identified with $\Q^n\subseteq(\A_{\Q})^n$
  and the result follows from the previous theorem.
\end{proof}

\begin{theorem}
  \lean{NumberField.AdeleRing.discrete}
  \label{NumberField.AdeleRing.discrete}
  \uses{NumberField.AdeleRing.zero_discrete}
  \leanok
  The additive subgroup $K$ of $\A_K$ is discrete.
\end{theorem}
\begin{proof}
  If $x\in K$ and $U$ is the open subset in the previous lemma, then
  it's easily checked that $K\cap U=\{0\}$ implies $K\cap (U+x)=\{x\}$,
  and $U+x$ is open.
\end{proof}

For compactness we follow the same approach.

\begin{theorem}
  \lean{Rat.AdeleRing.cocompact}
  \label{Rat.AdeleRing.cocompact}
  \leanok
  The quotient $\A_{\Q}/\Q$ is compact.
\end{theorem}
\begin{proof}
  The space $\prod_p\Z_p\times[0,1]\subseteq\A_{\Q}$ is a product of compact spaces
  and is hence compact. I claim that it surjects onto $\A_{\Q}/\Q$. Indeed,
  if $a\in\A_{\Q}$ then for the finitely many prime numbers $p\in S$ such that $a_p\not\in\Z_p$
  we have $a_p\in\frac{r_p}{p^{n_p}}+\Z_p$ with $r_p/p^{n_p}\in\Q$, and
  if $q=\sum_{p\in S}\frac{r_p}{p^{n_p}}\in\Q$ then $a-q\in \prod_p\Z_p\times\R$.
  Now just subtract $\lfloor a_{\infty}-q\rfloor$ to move into $\prod_p\Z_p\times[0,1)$
  and we are done.
\end{proof}

\begin{theorem}
  \lean{NumberField.AdeleRing.cocompact}
  \label{NumberField.AdeleRing.cocompact}
  \uses{Rat.AdeleRing.cocompact,NumberField.AdeleRing.baseChangeEquiv}
  \leanok
  The quotient $\A_K/K$ is compact.
\end{theorem}
\begin{proof}
  We proceed as in the discreteness proof above, by reducing to $\Q$. As before, choosing
  a $\Q$-basis of $K$ gives us $\A_K/K\cong(\A_{\Q}/\Q)^n$ so the result follows from
  the previous theorem.
\end{proof}
